---
title: "A generated allocation fit and capacity scenario"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A generated allocation fit and capacity scenario}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This workflow is experimental. Its model family is corrected static CLM with an
explicit outside option; a broader recipe interface is not part of this example.

For model selection and the existing public FCA interface, begin with
[Choosing an FCA index or an allocation model](model-choice.html). The
[preparation guide](spax-102-data-prep.html) explains counts, units, IDs and
missing travel; [saved-result inspection](function-naming.html) describes the
standard R methods used below. An FCA accessibility index and CLM contact
probability retain their different meanings.

The first section is a small fit and prediction example. The second section
executes the complete 399-attempt facility bootstrap, including scenario
evaluation, tables, plots and a fresh installed R process. It is intentionally
more expensive and needs no network connection. Allow it to finish before
starting another resource-intensive validation; shortening the attempt count
would not demonstrate the supported interval procedure.

This example allocates mutually exclusive service requests to facilities or an
outside category during one reference period. Demand is request opportunities,
supply is service slots, and travel is minutes. Supply affects attraction and
proportional capacity attribution in this toy model. Attributed capacity is
neither delivered service nor a hard limit on facility utilization. The outside
category is not automatically unmet need.

```{r setup}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spax)
origin_ids <- paste0("o", 1:8)
facility_ids <- paste0("f", 1:4)
demand <- setNames(c(1200, 800, 1000, 900, 1500, 1100, 0, NA), origin_ids)
supply <- setNames(c(80, 60, 90, 70), facility_ids)
travel <- outer(c(2, 7, 13, 18, 23, 29, 10, 21), c(3, 11, 20, 28),
                function(origin, facility) abs(origin - facility) + 2)
dimnames(travel) <- list(origin_ids, facility_ids)
units <- list(demand_units = "requests", demand_period = "one_reference_period",
              supply_units = "service_slots", supply_period = "one_reference_period",
              travel_units = "minutes")
```

For a reproducible demonstration, generate rounded expected facility counts
using a declared travel scale and outside weight. These are synthetic counts,
not empirical evidence for any travel family or feedback mechanism.

```{r generated-counts}
active <- which(is.finite(demand) & demand > 0)
weights <- sweep(exp(-travel[active, ]^2 / (2 * 9^2)), 2, supply, "*")
probability <- weights / (rowSums(weights) + 30)
observed <- setNames(round(colSums(probability * demand[active])), facility_ids)
observed <- observed[c("f3", "f1", "f4", "f2")] # names determine alignment
```

Preparation records positive, zero and unknown demand separately. Only finite
positive demand contributes to fitted facility loads. Strict numeric preparation
rejects unknown active travel unless its policy is explicitly changed.

```{r prepare-and-fit}
prepared <- prepare_allocation(demand, supply, travel,
  missing_demand = "exclude", metadata = units)
model <- clm_allocation(prepared, family = "gaussian", fit_v0 = TRUE)
starts <- rbind(near = c(sigma = 8, v0 = 25),
                short = c(sigma = 5, v0 = 60),
                long = c(sigma = 15, v0 = 15))
fit <- fit_allocation(model, observed, starts,
  lower = c(sigma = 1, v0 = 1), upper = c(sigma = 30, v0 = 150),
  output = "utilization", loss = "poisson", gradient = TRUE,
  control = list(maxit = 150))
summary(fit)
coef(fit)
fitted(fit)
```

The output includes eligible and failed starts, boundaries, messages and
disagreement among the supplied starts. Agreement is a numerical diagnostic,
not proof of a global optimum or parameter identification. Choosing expected
counts as the target does not choose a likelihood; Poisson is explicit here.
Joint Poisson fitting does not impose exact total matching.

Evaluate at the selected parameters, then request predictions for every origin.
Requested demand multiplies predicted probabilities into expected counts. It
does not change the saved facility loads. Origin o7 has known zero demand;
o8 has unknown demand. Their contact probabilities may still be available.

```{r evaluate-and-predict}
baseline <- evaluate_allocation(model, theta = coef(fit))
baseline$coverage$regional
prediction <- predict_allocation(baseline, travel, demand = demand,
  outputs = c("rho", "A", "abar", "flow"))
prediction$validity
data.frame(origin_id = origin_ids, rho = prediction$outputs$rho,
           A = prediction$outputs$A, abar = prediction$outputs$abar)
prediction$outputs$flow
```

Probability validity, demand availability and capacity attribution have
different conditions. Unknown travel can invalidate a normalized choice row.
A facility with positive supply and zero saved load can have valid contact
probability at a new origin while its unqualified capacity attribution remains
unknown. The prediction result preserves these distinctions.

Here rho is the inside-contact probability. With facility load U and supply S,
A is the sum over facilities of P times S/U: attributed slots per request
opportunity. For rho above zero, abar=A/rho is attributed slots per inside-contact
request. These are capacity-attribution measures under the stated model; flow is
the expected request count D times P. Requested demand and travel must use the
saved reference period and units. The interface does not convert units.

Change one facility's capacity by preparing and evaluating a new system at the
same theta. This is a model-based fixed-parameter scenario, not an estimated
causal intervention effect. Changing supply updates attraction, allocation and
facility loads as well as proportional capacity attribution.

```{r capacity-scenario}
scenario_supply <- supply
scenario_supply["f2"] <- 1.2 * scenario_supply["f2"]
scenario_inputs <- prepare_allocation(demand, scenario_supply, travel,
  missing_demand = "exclude", metadata = units)
scenario_model <- clm_allocation(scenario_inputs, family = "gaussian", fit_v0 = TRUE)
scenario <- evaluate_allocation(scenario_model, theta = coef(fit))
scenario_prediction <- predict_allocation(scenario, travel, demand = demand,
  outputs = c("rho", "A", "abar"))
data.frame(origin_id = origin_ids,
  baseline_contact = prediction$outputs$rho,
  scenario_contact = scenario_prediction$outputs$rho,
  baseline_attribution = prediction$outputs$A,
  scenario_attribution = scenario_prediction$outputs$A)
```

The evaluated source contains a small prediction snapshot. Saving it does not
require retaining the original model closure or travel matrix. A fresh process
can read the saved source and predict from a new requested travel matrix.

```{r save-and-reload}
saved <- tempfile(fileext = ".rds")
saveRDS(scenario, saved)
restored <- readRDS(saved)
restored_prediction <- predict_allocation(restored, travel, demand = demand,
  outputs = c("rho", "A", "abar"))
stopifnot(identical(restored_prediction, scenario_prediction))
unlink(saved)
```

An optional map needs an explicit mapping from origin IDs to raster cells.
This synthetic grid has no geographic interpretation. Mapping is separate
from origin registry order, and map construction does not change load support.

```{r optional-map, fig.width=5, fig.height=3}
template <- terra::rast(nrows = 2, ncols = 4, xmin = 0, xmax = 4,
                        ymin = 0, ymax = 2, crs = "EPSG:3857")
spatial <- list(template = template,
  cell_index = setNames(c(6, 2, 8, 1, 7, 3, 4, 5), origin_ids))
contact_map <- predict_allocation(restored, travel, outputs = "rho",
  spatial = spatial, type = "raster")
terra::plot(contact_map)
```

## Facility-count bootstrap on a larger generated network

The four-facility example above explains preparation and prediction support. For
uncertainty, use a separate generated network of 80 origins and 40 facilities.
The observation unit is one facility total during the declared period. Here the
counts are generated independently from a Poisson law. Independence across
facilities is an explicit assumption; it does not follow merely from mutually
exclusive allocation destinations. In an application, that law needs its own
justification.

```{r bootstrap-generated-inputs}
i <- seq_len(80)
j <- seq_len(40)
boot_travel <- sqrt(
  outer(seq(2, 58, length.out = 80), seq(4, 56, length.out = 40), "-")^2 +
  outer(8 * sin(i * .73), 7 * sin(j * 1.31), "-")^2) + 2
dimnames(boot_travel) <- list(paste0("origin", i), paste0("facility", j))
boot_demand <- setNames(80 + 35 * (1 + sin(i * .23)), rownames(boot_travel))
boot_supply <- setNames(rep(c(5, 7, 6, 8, 9), 8), colnames(boot_travel))
truth <- c(sigma = 12, v0 = 45)
boot_score <- exp(-boot_travel^2 / (2 * truth[["sigma"]]^2)) *
  rep(boot_supply, each = length(boot_demand))
boot_probability <- boot_score / (rowSums(boot_score) + truth[["v0"]])
set.seed(2490001)
boot_observed <- setNames(rpois(40, colSums(boot_probability * boot_demand)),
  names(boot_supply))
boot_inputs <- prepare_allocation(boot_demand, boot_supply, boot_travel,
  metadata = units)
boot_model <- clm_allocation(boot_inputs, family = "gaussian", fit_v0 = TRUE)
boot_starts <- rbind(short = c(sigma = 8, v0 = 25),
  medium = c(sigma = 15, v0 = 70), wide = c(sigma = 25, v0 = 15))
boot_fit <- fit_allocation(boot_model, boot_observed, boot_starts,
  lower = c(sigma = 2, v0 = 1), upper = c(sigma = 60, v0 = 500),
  output = "utilization", loss = "poisson", gradient = TRUE,
  control = list(maxit = 300, factr = 10, pgtol = 1e-8))
as.data.frame(boot_fit)
summary(boot_fit)
```

The fit table contains saved parameter estimates. Diagnostic plotting delegates
to the selected fit and does not refit the model.

```{r selected-fit-plot, fig.width=5, fig.height=4}
plot(boot_fit, type = "fit")
```

Prepare a supply-only scenario that increases the second facility's capacity by
20%. Each bootstrap draw evaluates both baseline and changed supply at the same
parameters. This pairs the contrast while allowing attraction, allocation and
loads to respond to supply. It does not estimate a causal intervention effect.

```{r paired-bootstrap}
boot_scenario_supply <- boot_supply
boot_scenario_supply["facility2"] <- 1.2 * boot_scenario_supply["facility2"]
boot_scenario_inputs <- prepare_allocation(boot_demand, boot_scenario_supply,
  boot_travel, metadata = units)
boot_scenario <- clm_allocation(boot_scenario_inputs, family = "gaussian",
  fit_v0 = TRUE)
observation <- list(event = "allocated service requests",
  period = "one_reference_period", sampling = "independent_facilities")
boot <- bootstrap_allocation(boot_model, boot_fit, observation,
  scenario = boot_scenario, seed = 20260930)
summary(boot)
as.data.frame(boot)
head(as.data.frame(boot, type = "draws"))
head(as.data.frame(boot, type = "starts"))
```

The procedure makes 399 predetermined attempts. It resamples facility
observation contributions while keeping all facilities in the allocation
denominator. It retains unsuccessful attempts and successful inner-boundary
fits; it does not replace failed attempts until a desired success count is
reached. Inspect the actual and planned denominators and each saved quantity's
status. An eligible interval requires a complete run of all 399 attempts, at least
380 finite successful values and an eligible original fit. A refused original
may have zero attempted draws and
still preserve its point estimates and diagnostic reasons.

These are pointwise percentile bootstrap intervals conditional on the declared observation
assumption, fixed demand, supply, travel, included facilities and mean model.
Generated-study acceptance does not establish an application's observation law
or guarantee exact 95% coverage. Regional A is supported raw supply per retained
demand and has no sampling interval. Supply at facilities without a finite edge
on retained demand support is excluded from attributed capacity; the summary
displays total, attributed and unsupported supply separately. The grouped-total HC3 refusal
is unchanged; this facility bootstrap is a separate supported procedure.

```{r bootstrap-plot, fig.width=6, fig.height=4}
plot(boot, quantity = "regional_rho", type = "interval")
```

Coverage extraction returns saved regional or cell tables. The cells are the
retained positive-demand origins, with canonical origin IDs; extraction does not
expand prediction support or create spatial uncertainty intervals. Unit and
support declarations remain attached as attributes, so keep them when exporting
tables to a format that drops R attributes.

```{r saved-coverage-tables}
boot_evaluated <- evaluate_allocation(boot_model, coef(boot_fit))
regional_table <- as.data.frame(boot_evaluated$coverage)
cell_table <- as.data.frame(boot_evaluated$coverage, type = "cells")
regional_table
head(cell_table)
attr(cell_table, "input_metadata")
```

CSV files make these saved tables available to other software. They do not
retain R attributes or result classes. Keep an RDS bundle alongside them to
preserve observation declarations, units, support, reporting settings and the
complete saved results. The estimates and draws include `delta_contacts`, the
paired change in expected inside contacts under the supply scenario.

```{r export-saved-results}
export_dir <- tempfile("allocation-results-")
dir.create(export_dir)
export_tables <- list(estimates = as.data.frame(boot),
  draws = as.data.frame(boot, type = "draws"),
  starts = as.data.frame(boot, type = "starts"),
  parameters = as.data.frame(boot_fit, type = "parameters"),
  regional = regional_table, cells = cell_table)
for (name in names(export_tables)) {
  write.csv(export_tables[[name]], file.path(export_dir, paste0(name, ".csv")),
    row.names = FALSE)
}
bundle_path <- file.path(export_dir, "allocation-results.rds")
saveRDS(list(bootstrap = boot, fit = boot_fit, coverage = boot_evaluated$coverage,
  tables = export_tables,
  metadata = list(observation = boot$observation,
    input_metadata = attr(cell_table, "input_metadata"),
    support = attr(cell_table, "support"),
    reporting_settings = attr(cell_table, "reporting_settings"))), bundle_path)
list.files(export_dir)
as.data.frame(boot)[as.data.frame(boot)$quantity == "delta_contacts", ]
```

The optional map remains a point prediction. This artificial 8-by-10 grid has no
geographic meaning; a bootstrap of regional quantities does not provide cellwise
or simultaneous map intervals.

```{r generated-point-map, fig.width=5, fig.height=4}
boot_template <- terra::rast(nrows = 8, ncols = 10, xmin = 0, xmax = 10,
  ymin = 0, ymax = 8, crs = "EPSG:3857")
boot_spatial <- list(template = boot_template,
  cell_index = setNames(seq_len(80), names(boot_demand)))
boot_contact_map <- predict_allocation(boot_evaluated, boot_travel,
  outputs = "rho", spatial = boot_spatial, type = "raster")
terra::plot(boot_contact_map, main = "Generated point contact probability")
```

Finally, inspect the saved results in a new R process using the installed
package. This check reads saved tables and summaries without solving or drawing
another bootstrap sample, and exercises printing and plotting after reload.

```{r fresh-installed-reload}
child_path <- tempfile(fileext = ".R")
literal <- function(x) paste(capture.output(dput(x)), collapse = "\n")
writeLines(c(
  paste0(".libPaths(c(", literal(dirname(find.package("spax"))), ", .libPaths()))"),
  "library(spax)",
  paste0("saved <- readRDS(", literal(bundle_path), ")"),
  "stopifnot(identical(as.data.frame(saved$bootstrap), saved$tables$estimates))",
  "stopifnot(identical(as.data.frame(saved$bootstrap, type = 'draws'), saved$tables$draws))",
  "stopifnot(identical(as.data.frame(saved$bootstrap, type = 'starts'), saved$tables$starts))",
  "stopifnot(identical(as.data.frame(saved$fit), saved$tables$parameters))",
  "stopifnot(identical(as.data.frame(saved$coverage), saved$tables$regional))",
  "stopifnot(identical(as.data.frame(saved$coverage, type = 'cells'), saved$tables$cells))",
  "stopifnot(identical(saved$bootstrap$observation, saved$metadata$observation))",
  "stopifnot(identical(attr(saved$tables$cells, 'input_metadata'), saved$metadata$input_metadata))",
  "stopifnot(inherits(summary(saved$bootstrap), 'summary.ae_allocation_bootstrap'))",
  "stopifnot(inherits(summary(saved$fit), 'summary.ae_multistart_fit'))",
  "print(saved$bootstrap)",
  "print(summary(saved$bootstrap))",
  "print(summary(saved$fit))",
  "print(as.data.frame(saved$bootstrap))",
  "plot_path <- tempfile(fileext = '.pdf')",
  "grDevices::pdf(plot_path)",
  "plot(saved$bootstrap, quantity = 'regional_rho')",
  "plot(saved$bootstrap, quantity = 'regional_rho', type = 'draws')",
  "plot(saved$fit, type = 'fit')",
  "grDevices::dev.off()",
  "stopifnot(file.info(plot_path)$size > 0)",
  "unlink(plot_path)",
  "cat('Fresh installed saved-result inspection passed\\n')"
), child_path)
rscript <- file.path(R.home("bin"), if (.Platform$OS.type == "windows") "Rscript.exe" else "Rscript")
status <- system2(rscript, c("--vanilla", shQuote(child_path)))
stopifnot(status == 0L)
unlink(child_path)
```
