---
title: "A generated allocation fit and capacity scenario"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A generated allocation fit and capacity scenario}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This workflow is experimental. Its model family is corrected static CLM with an
explicit outside option; a broader recipe interface is not part of this example.

This example allocates mutually exclusive service requests to facilities or an
outside category during one reference period. Demand is request opportunities,
supply is service slots, and travel is minutes. Supply affects attraction and
proportional capacity attribution in this toy model. Attributed capacity is
neither delivered service nor a hard limit on facility utilization. The outside
category is not automatically unmet need.

```{r setup}
library(spax)
origin_ids <- paste0("o", 1:8)
facility_ids <- paste0("f", 1:4)
demand <- setNames(c(1200, 800, 1000, 900, 1500, 1100, 0, NA), origin_ids)
supply <- setNames(c(80, 60, 90, 70), facility_ids)
travel <- outer(c(2, 7, 13, 18, 23, 29, 10, 21), c(3, 11, 20, 28),
                function(origin, facility) abs(origin - facility) + 2)
dimnames(travel) <- list(origin_ids, facility_ids)
units <- list(demand_units = "requests", demand_period = "one_reference_period",
              supply_units = "service_slots", supply_period = "one_reference_period",
              travel_units = "minutes")
```

For a reproducible demonstration, generate rounded expected facility counts
using a declared travel scale and outside weight. These are synthetic counts,
not empirical evidence for any travel family or feedback mechanism.

```{r generated-counts}
active <- which(is.finite(demand) & demand > 0)
weights <- sweep(exp(-travel[active, ]^2 / (2 * 9^2)), 2, supply, "*")
probability <- weights / (rowSums(weights) + 30)
observed <- setNames(round(colSums(probability * demand[active])), facility_ids)
observed <- observed[c("f3", "f1", "f4", "f2")] # names determine alignment
```

Preparation records positive, zero and unknown demand separately. Only finite
positive demand contributes to fitted facility loads. Strict numeric preparation
rejects unknown active travel unless its policy is explicitly changed.

```{r prepare-and-fit}
prepared <- prepare_allocation(demand, supply, travel,
  missing_demand = "exclude", metadata = units)
model <- clm_allocation(prepared, family = "gaussian", fit_v0 = TRUE)
starts <- rbind(near = c(sigma = 8, v0 = 25),
                short = c(sigma = 5, v0 = 60),
                long = c(sigma = 15, v0 = 15))
fit <- fit_allocation(model, observed, starts,
  lower = c(sigma = 1, v0 = 1), upper = c(sigma = 30, v0 = 150),
  output = "utilization", loss = "poisson", gradient = TRUE,
  control = list(maxit = 150))
summary(fit)
coef(fit)
fitted(fit)
```

The output includes eligible and failed starts, boundaries, messages and
disagreement among the supplied starts. Agreement is a numerical diagnostic,
not proof of a global optimum or parameter identification. Choosing expected
counts as the target does not choose a likelihood; Poisson is explicit here.
Joint Poisson fitting does not impose exact total matching.

Evaluate at the selected parameters, then request predictions for every origin.
Requested demand multiplies predicted probabilities into expected counts. It
does not change the saved facility loads. Origin o7 has known zero demand;
o8 has unknown demand. Their contact probabilities may still be available.

```{r evaluate-and-predict}
baseline <- evaluate_allocation(model, theta = coef(fit))
baseline$coverage$regional
prediction <- predict_allocation(baseline, travel, demand = demand,
  outputs = c("rho", "A", "abar", "flow"))
prediction$validity
data.frame(origin_id = origin_ids, rho = prediction$outputs$rho,
           A = prediction$outputs$A, abar = prediction$outputs$abar)
prediction$outputs$flow
```

Probability validity, demand availability and capacity attribution have
different conditions. Unknown travel can invalidate a normalized choice row.
A facility with positive supply and zero saved load can have valid contact
probability at a new origin while its unqualified capacity attribution remains
unknown. The prediction result preserves these distinctions.

Here rho is the inside-contact probability. With facility load U and supply S,
A is the sum over facilities of P times S/U: attributed slots per request
opportunity. For rho above zero, abar=A/rho is attributed slots per inside-contact
request. These are capacity-attribution measures under the stated model; flow is
the expected request count D times P. Requested demand and travel must use the
saved reference period and units. The interface does not convert units.

Change one facility's capacity by preparing and evaluating a new system at the
same theta. This is a model-based fixed-parameter scenario, not an estimated
causal intervention effect. Changing supply updates attraction, allocation and
facility loads as well as proportional capacity attribution.

```{r capacity-scenario}
scenario_supply <- supply
scenario_supply["f2"] <- 1.2 * scenario_supply["f2"]
scenario_inputs <- prepare_allocation(demand, scenario_supply, travel,
  missing_demand = "exclude", metadata = units)
scenario_model <- clm_allocation(scenario_inputs, family = "gaussian", fit_v0 = TRUE)
scenario <- evaluate_allocation(scenario_model, theta = coef(fit))
scenario_prediction <- predict_allocation(scenario, travel, demand = demand,
  outputs = c("rho", "A", "abar"))
data.frame(origin_id = origin_ids,
  baseline_contact = prediction$outputs$rho,
  scenario_contact = scenario_prediction$outputs$rho,
  baseline_attribution = prediction$outputs$A,
  scenario_attribution = scenario_prediction$outputs$A)
```

The evaluated source contains a small prediction snapshot. Saving it does not
require retaining the original model closure or travel matrix. A fresh process
can read the saved source and predict from a new requested travel matrix.

```{r save-and-reload}
saved <- tempfile(fileext = ".rds")
saveRDS(scenario, saved)
restored <- readRDS(saved)
restored_prediction <- predict_allocation(restored, travel, demand = demand,
  outputs = c("rho", "A", "abar"))
stopifnot(identical(restored_prediction, scenario_prediction))
unlink(saved)
```

An optional map needs an explicit mapping from origin IDs to raster cells.
This synthetic grid has no geographic interpretation. Mapping is separate
from origin registry order, and map construction does not change load support.

```{r optional-map, fig.width=5, fig.height=3}
template <- terra::rast(nrows = 2, ncols = 4, xmin = 0, xmax = 4,
                        ymin = 0, ymax = 2, crs = "EPSG:3857")
spatial <- list(template = template,
  cell_index = setNames(c(6, 2, 8, 1, 7, 3, 4, 5), origin_ids))
contact_map <- predict_allocation(restored, travel, outputs = "rho",
  spatial = spatial, type = "raster")
terra::plot(contact_map)
```
