---
title: "spax Package Function Naming Convention"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spax Package Function Naming Convention}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spax)
```

# SPAX Package Development Guidelines

## Core Principles

1.  **Modularity**: Functions should be single-purpose and composable
2.  **Consistency**: Follow established naming and structure conventions
3.  **Documentation**: Provide clear, example-driven documentation
4.  **Performance**: Consider computational efficiency
5.  **Testing**: Comprehensive test coverage

## Function Hierarchy

### 1. High-Level Functions (`spax_*`)

-   **Purpose**: Complete workflows for direct user interaction

-   **Naming**: Must start with `spax_`

-   **Examples**:

    ``` r
    spax_e2sfca(demand, supply, distance)
    spax_ifca(distance_raster, demand, supply, sigma)
    ```

### 2. Semi-High Level Functions (`compute_*`)

-   **Purpose**: Complex computational operations
-   **Naming**: Must start with `compute_`
-   **Examples**: `compute_iterative()`, `compute_iterative_fast()`

### 3. Middle-Level Operations

Four categories of operational functions:

#### a. Gather Operations (`gather_*`)

-   **Purpose**: Dimension reduction (raster stack → vector)
-   **Examples**: `gather_demand()`, `gather_weighted()`

#### b. Spread Operations (`spread_*`)

-   **Purpose**: Dimension expansion (vector → raster stack)
-   **Examples**: `spread_access()`, `spread_weighted()`

#### c. Calculate Operations (`calc_*`)

-   **Purpose**: Layer-wise operations maintaining dimensions
-   **Examples**: `calc_decay()`, `calc_normalize()`

#### d. Transform Operations (`transform_*`)

-   **Purpose**: Surface-wide transformations
-   **Examples**: `transform_pmf()`

#### e. Independent function (`sample_*`)

### 4. Low-Level Functions (Internal)

-   **Validation** (`.chck_*`): Input validation
-   **Helpers** (`.help_*`): Utility functions
-   **Core** (`.*_core`): Main computational logic

## Function Structure

### Standard Template

Note that not all functions will have all sections. Adjust as needed.

``` r
function_name <- function(required_param,
                         optional_param = default_value,
                         ...,
                         snap = FALSE) {
    # 1. Validation
    if (!snap) {
        .chck_function_name(required_param, optional_param)
    }
    
    # 2. Pre-processing
    processed_data <- .help_process_data(required_param)
    
    # 3. Core Computation
    result <- .function_name_core(processed_data, ...)
    
    # 4. Post-processing
    if (!snap) {
        result <- .help_post_process(result)
    }
    
    return(result)
}
```

## Examples and Data Usage

### Using Package Data

When creating examples, prefer using package datasets from `data.R`:

1.  **Healthcare Facilities** + **Distance raster/Isochrone**: `hc12_phc` and `hc12_hos`+ `hos_iscr` and `phc_iscr`

    ``` r
    # Example using hospital data
    weights <- calc_decay(hos_iscr,
                         method = "gaussian",
                         sigma = 30)

    choice_probs <- calc_choice(weights,
                               attractiveness = hc12_hos$s_doc)
    ```

2.  **Population Data**: `u5pd` (Under-5 Population Density)

    ``` r
    # Example using population data
    pmf_result <- transform_pmf(u5pd, return_total = TRUE)
    ```

3.  **Administrative Boundaries**: `bound0`, `bound1`

    ``` r
    # Example using boundary data
    plot(bound1) +
      geom_sf(aes(fill = ADM1_EN))
    ```

### Creating New Examples

If package data isn't suitable, create minimal reproducible examples:

``` r
# Create test raster
r <- terra::rast(nrows = 10, ncols = 10)
values(r) <- runif(100)

# Create test points
pts <- data.frame(
    x = runif(5, 0, 1),
    y = runif(5, 0, 1),
    capacity = rpois(5, 10)
)
```

## Documentation Requirements

### Function Documentation

``` r
#' @title Function Title
#' @description Detailed description
#' @param param1 Description of first parameter
#' @param snap Logical; if TRUE enables fast computation mode
#' @examples
#' # Use package data when possible
#' data(hc12_hos)
#' result <- function_name(hc12_hos$s_doc)
#' 
#' # Or create minimal example
#' x <- rnorm(10)
#' result <- function_name(x)
#' @export
```

### Internal Function Documentation

``` r
#' @keywords internal
.internal_function <- function() {
    # Implementation
}
```

## File Organization

### File Naming

`{level}-{description}.R`:

\- `01-*`: Core building blocks

\- `02-*`: Assembly-level functions

\- `03-*`: Main workflows

\- `09-*`: Independent utilities

### Location

All R files must be in the `R/` directory (no subdirectories)

## Testing Guidelines

### Test Organization

``` r
# tests/testthat/test-function_name.R

test_that("function handles basic cases", {
    # Use package data - only when nescessary (since it's big)
    result <- function_name(hc12_hos$s_doc)
    expect_equal(...)
    
    # Or create test data
    x <- 1:10
    result <- function_name(x)
    expect_equal(...)
})
```

### Test Coverage

-   Normal operation
-   Edge cases
-   Error conditions
-   `snap = TRUE` mode
-   Performance benchmarks for critical functions

## Version Control

### Commit Messages

```         
type(scope): description

- feat(calc): add new decay function
- fix(normalize): handle NA values correctly
- docs(examples): update with package data
```

### Branches

-   `main`: Stable release
-   `develop`: Development version
-   `feature/*`: New features
-   `fix/*`: Bug fixes

## Performance Considerations

1.  Include `snap` parameter for functions called repeatedly
2.  Use `terra` functions instead of `raster`
3.  Vectorize operations when possible
4.  Consider parallel processing for intensive operations

## Documentation Updates

When modifying functions:

1\. Update relevant vignettes

2\. Update `README.md` if needed

3\. Run roxygen2

4\. Update `NEWS.md`

5\. Update examples to use package data where appropriate
