---
title: "Finding functions and inspecting results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding functions and inspecting results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spax)
```

spax retains its public FCA functions alongside the experimental static
allocation workflow. Names describe the interface they belong to; there is no
rule that every complete workflow must start with `spax_`. Existing FCA calls
and result types remain supported.

## Choose the level of operation

| Task | Public interface | What it returns |
|:--|:--|:--|
| Binary catchment FCA | `spax_2sfca()` | `spax` result with accessibility raster |
| Distance-weighted FCA | `spax_e2sfca()` | `spax` result with accessibility raster |
| Squared access-side decay FCA | `spax_m2sfca()` | `spax` result with accessibility raster |
| Supply/demand calculation using supplied kernels | `compute_access()` | Accessibility `SpatRaster` |
| Decay, normalization or choice weights | `calc_decay()`, `calc_normalize()`, `calc_choice()` | See the individual input/output contracts |
| Weighted reduction to facilities | `gather_weighted()`, `gather_demand()` | Facility vector/table |
| Weighted distribution to origins | `spread_weighted()`, `spread_access()` | Raster result |
| Checked allocation preparation and declaration | `prepare_allocation()`, `clm_allocation()` | Prepared inputs and experimental model |
| Fit, evaluate, predict and bootstrap | `fit_allocation()`, `evaluate_allocation()`, `predict_allocation()`, `bootstrap_allocation()` | Model-specific saved results |

These function families share concepts such as supply, demand and facility IDs.
Their arguments and result meanings are not interchangeable. In particular,
`calc_normalize()` applied to a numeric matrix normalizes the whole matrix,
whereas a multilayer raster is normalized across layers at each cell. Select the
operation appropriate to the declared axes; do not infer behavior from a name.

```{r public-interface-check}
functions <- c("spax_2sfca", "spax_e2sfca", "spax_m2sfca", "compute_access",
  "prepare_allocation", "clm_allocation", "fit_allocation",
  "evaluate_allocation", "predict_allocation", "bootstrap_allocation")
stopifnot(all(functions %in% getNamespaceExports("spax")))
data.frame(function_name = functions)
```

The former iterative-FCA interface is no longer available. It is not an alias
for static CLM or a supported fixed-point workflow. Private functions beginning
with `.` and other unexported implementation functions are not public contracts;
examples for users should not call them with `:::`.

## Inspect saved results using standard R generics

| Result | Inspection | Interpretation |
|:--|:--|:--|
| FCA `spax` object | `print()`, `summary()`, `plot()`, `$accessibility` | Saved supply-per-demand index; raster-cell summary |
| Allocation multistart fit | `coef()`, `fitted()`, `summary()`, `as.data.frame(type = "parameters")`, `plot()` | Selected fit plus retained per-start diagnostics |
| Allocation coverage | `summary()`, `as.data.frame(type = "regional")` or `type = "cells"` | Saved coverage on retained positive-demand support |
| Allocation bootstrap | `summary()`, `as.data.frame(type = "estimates")`, `"draws"`, `"starts"`, `plot()` | Saved point/interval status and every attempted result |

Inspection does not refit the allocation model or resample the bootstrap. A
failed search can still be summarized, but parameter extraction requires an
eligible selected fit. Bootstrap methods preserve incomplete and refused
statuses; displaying a point estimate does not make an interval available.

FCA rasters and allocation numeric tables have different storage contracts. Use
`terra::as.data.frame()` to extract raster cell values, and preserve geometry
separately. Allocation tables carry IDs; coverage tables also carry unit,
support and reporting attributes. CSV loses those attributes, so retain the
documented RDS bundle when exporting allocation results.

## Keep input validation enabled

Some legacy functions expose `snap`. Leave its default `FALSE` in ordinary user
code. It is an internal optimization switch, not a different model or a general
speed recommendation. New prepared allocation inputs validate IDs, geometry and
support at their own boundary and do not use a universal `snap` convention.

See [Choosing a model](model-choice.html) for a runnable comparison,
[the FCA calculation](spax-103-dissection.html) for public building blocks, and
[the complete allocation workflow](allocation-workflow.html) for fitting,
prediction, intervals, scenarios and persistence.
