---
title: "spax Package Function Naming Convention"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spax Package Function Naming Convention}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spax)
```

# SPAX Package Development Guidelines

## Function Hierarchy and Naming Conventions

### High-Level Functions (`spax_*`)

-   **Purpose**: Complete workflows that users interact with directly
-   **Naming**: Must start with `spax_`
-   **Examples**: `spax_e2sfca()`, `spax_ifca()`
-   **Input Requirements**: Should accept primitive inputs (demand, supply, distance rasters)
-   **Documentation**: Must include complete examples in roxygen comments

### Semi-High Level Functions (`comp_*`)

-   **Purpose**: Complex computational operations that can work independently
-   **Naming**: Must start with `comp_`
-   **Examples**: `comp_iterative()`, `comp_iterative_fast()`
-   **Note**: These functions often implement core algorithms

### Middle-Level Operations

Four categories of operational functions:

1.  **Gather Operations** (`gather_*`)
    -   Purpose: Dimension reduction (e.g., raster stack → vector)
    -   Examples: `gather_demand()`, `gather_weighted()`
2.  **Spread Operations** (`spread_*`)
    -   Purpose: Dimension expansion (e.g., vector → raster stack)
    -   Examples: `spread_access()`, `spread_weighted()`
3.  **Calculate Operations** (`calc_*`)
    -   Purpose: Layer-wise operations maintaining dimensions
    -   Examples: `calc_decay()`, `calc_normalize()`, `calc_choice()`
4.  **Transform Operations** (`transform_*`)
    -   Purpose: Surface-wide transformations
    -   Examples: `transform_pmf()`

### Low-Level Functions

Internal functions not exported in the NAMESPACE:

1.  **Validation Functions** (`.chck_*`)
    -   Purpose: Input validation
    -   Examples: `.chck_calc_choice()`, `.chck_sample_pmf()`
2.  **Helper Functions** (`.help_*`)
    -   Purpose: Utility functions used by higher-level functions
    -   Examples: `.help_prep_facilities()`, `.help_gen_sample_size()`
3.  **Core Functions** (`.*_core`)
    -   Purpose: Main computational logic for complex operations
    -   Examples: `.sample_pmf_core_indp()`, `.sample_pmf_core_evlv()`

## Function Structure Guidelines

### Standard Function Template

``` r
function_name <- function(
    required_param,
    optional_param = default_value,
    ...,
    snap = FALSE
) {
    # 1. Input Validation (skip if snap = TRUE)
    if (!snap) {
        .chck_function_name(required_param, optional_param)
    }
    
    # 2. Pre-processing
    processed_data <- .help_process_data(required_param)
    
    # 3. Main Computation
    result <- .function_name_core(processed_data, ...)
    
    # 4. Post-processing (skip if snap = TRUE)
    if (!snap) {
        result <- .help_post_process(result)
    }
    
    return(result)
}
```

### The `snap` Parameter

-   **Purpose**: Performance optimization for internal use
-   **Default**: Always `FALSE`
-   **When TRUE**:
    -   Skips input validation
    -   Skips non-essential processing
    -   Assumes inputs are correctly formatted
    -   May skip naming/attribute assignment
-   **Usage**: Only set `TRUE` when called from within other package functions where inputs have already been validated

### Documentation Requirements

1.  **Roxygen Headers**

``` r
#' @title Function Title
#' @description Detailed description
#' @param snap Logical; if TRUE enables fast computation mode with minimal validation
#' @export
```

2.  **Internal Function Documentation**

``` r
#' @keywords internal
.internal_function <- function() {
```

### File Organization

All R files must be in the `R/` directory (no subdirectories). Use file naming conventions:

-   `00-utils.R`: Helper functions
-   `01-core.R`: Core computations
-   `02-operations.R`: Middle-level operations
-   `03-compute.R`: Semi-high level functions
-   `04-spax.R`: High-level functions

### Error Handling

1.  Use `stop()` for critical errors in validation
2.  Use `warning()` for non-critical issues
3.  Use `message()` sparingly for important information

### Performance Considerations

1.  Always include the `snap` parameter for functions that might be called repeatedly
2.  Use `terra` functions instead of `raster` when possible
3.  Avoid loops when vectorized operations are possible
4.  Consider parallel processing for computationally intensive operations

### Testing Requirements

1.  Each function should have corresponding tests in `tests/testthat/`
2.  Include tests for both normal and `snap=TRUE` modes
3.  Test error conditions and edge cases
4.  Include performance benchmarks for critical functions

## Development Workflow

1.  **Creating New Functions**
    -   Start with the function template
    -   Add validation function if needed
    -   Implement core logic
    -   Add documentation
    -   Add tests
2.  **Modifying Existing Functions**
    -   Maintain backward compatibility
    -   Update documentation
    -   Update tests
    -   Update vignettes if behavior changes
3.  **Deprecating Functions**
    -   Use `.Deprecated()` function
    -   Maintain old function for at least one version
    -   Update documentation to point to new function
    -   Update vignettes to use new function

## Version Control Guidelines

1.  **Commit Messages**
    -   Start with type: feat, fix, docs, style, refactor, test, chore
    -   Include scope in parentheses
    -   Example: `feat(calc): add new decay function`
2.  **Branches**
    -   `main`: Stable release
    -   `develop`: Development version
    -   `feature/*`: New features
    -   `fix/*`: Bug fixes

## Documentation Updates

When adding or modifying functions:

1\. Update relevant vignettes

2\. Update README.md if needed

3\. Update NAMESPACE

4\. Run roxygen2 to update documentation

5\. Update NEWS.md with changes

## Missing Functions (Suggested Additions)

| Function Name               | Level  | Subcategory | Purpose                      |
|--------------------|---------------|---------------|---------------------|
| `spax_2sfca()`              | High   | Main        | Complete 2SFCA workflow      |
| `spax_montecarlo()`         | High   | Main        | Monte Carlo analysis         |
| `summarize_accessibility()` | Middle | Summarize   | Accessibility statistics     |
| `validate_inputs_check()`   | Low    | Check       | General input validation     |
| `plot_accessibility()`      | Middle | Visualize   | Standard accessibility plots |

Would you like to: 1. Review specific function groups in detail? 2. Discuss the suggested name changes? 3. Prioritize which missing functions to implement first?
