---
title: "spax Package Function Naming Convention"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spax Package Function Naming Convention}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spax)
```

# SPAX Package Development Guidelines

## Function Hierarchy and Naming Conventions

### High-Level Functions (`spax_*`)

-   **Purpose**: Complete workflows that users interact with directly
-   **Naming**: Must start with `spax_`
-   **Examples**: `spax_e2sfca()`, `spax_ifca()`
-   **Input Requirements**: Should accept primitive inputs (demand, supply, distance rasters)
-   **Documentation**: Must include complete examples in roxygen comments

### Semi-High Level Functions (`comp_*`)

-   **Purpose**: Complex computational operations that can work independently
-   **Naming**: Must start with `comp_`
-   **Examples**: `comp_iterative()`, `comp_iterative_fast()`
-   **Note**: These functions often implement core algorithms

### Middle-Level Operations

Four categories of operational functions:

1.  **Gather Operations** (`gather_*`)
    -   Purpose: Dimension reduction (e.g., raster stack → vector)
    -   Examples: `gather_demand()`, `gather_weighted()`
2.  **Spread Operations** (`spread_*`)
    -   Purpose: Dimension expansion (e.g., vector → raster stack)
    -   Examples: `spread_access()`, `spread_weighted()`
3.  **Calculate Operations** (`calc_*`)
    -   Purpose: Layer-wise operations maintaining dimensions
    -   Examples: `calc_decay()`, `calc_normalize()`, `calc_choice()`
4.  **Transform Operations** (`transform_*`)
    -   Purpose: Surface-wide transformations
    -   Examples: `transform_pmf()`

### Low-Level Functions

Internal functions not exported in the NAMESPACE:

1.  **Validation Functions** (`.chck_*`)
    -   Purpose: Input validation
    -   Examples: `.chck_calc_choice()`, `.chck_sample_pmf()`
2.  **Helper Functions** (`.help_*`)
    -   Purpose: Utility functions used by higher-level functions
    -   Examples: `.help_prep_facilities()`, `.help_gen_sample_size()`
3.  **Core Functions** (`.*_core`)
    -   Purpose: Main computational logic for complex operations
    -   Examples: `.sample_pmf_core_indp()`, `.sample_pmf_core_evlv()`

## Function Structure Guidelines

### Standard Function Template

``` r
function_name <- function(
    required_param,
    optional_param = default_value,
    ...,
    snap = FALSE
) {
    # 1. Input Validation (skip if snap = TRUE)
    if (!snap) {
        .chck_function_name(required_param, optional_param)
    }
    
    # 2. Pre-processing
    processed_data <- .help_process_data(required_param)
    
    # 3. Main Computation
    result <- .function_name_core(processed_data, ...)
    
    # 4. Post-processing (skip if snap = TRUE)
    if (!snap) {
        result <- .help_post_process(result)
    }
    
    return(result)
}
```

### The `snap` Parameter

-   **Purpose**: Performance optimization for internal use
-   **Default**: Always `FALSE`
-   **When TRUE**:
    -   Skips input validation
    -   Skips non-essential processing
    -   Assumes inputs are correctly formatted
    -   May skip naming/attribute assignment
-   **Usage**: Only set `TRUE` when called from within other package functions where inputs have already been validated

### Documentation Requirements

1.  **Roxygen Headers**

``` r
#' @title Function Title
#' @description Detailed description
#' @param snap Logical; if TRUE enables fast computation mode with minimal validation
#' @export
```

2.  **Internal Function Documentation**

``` r
#' @keywords internal
.internal_function <- function() {
```

### File Organization

All R files must be in the `R/` directory (no subdirectories). Use file naming conventions:

-   `00-utils.R`: Helper functions
-   `01-core.R`: Core computations
-   `02-operations.R`: Middle-level operations
-   `03-compute.R`: Semi-high level functions
-   `04-spax.R`: High-level functions

### Error Handling

1.  Use `stop()` for critical errors in validation
2.  Use `warning()` for non-critical issues
3.  Use `message()` sparingly for important information

### Performance Considerations

1.  Always include the `snap` parameter for functions that might be called repeatedly
2.  Use `terra` functions instead of `raster` when possible
3.  Avoid loops when vectorized operations are possible
4.  Consider parallel processing for computationally intensive operations

### Testing Requirements

1.  Each function should have corresponding tests in `tests/testthat/`
2.  Include tests for both normal and `snap=TRUE` modes
3.  Test error conditions and edge cases
4.  Include performance benchmarks for critical functions

## Development Workflow

1.  **Creating New Functions**
    -   Start with the function template
    -   Add validation function if needed
    -   Implement core logic
    -   Add documentation
    -   Add tests
2.  **Modifying Existing Functions**
    -   Maintain backward compatibility
    -   Update documentation
    -   Update tests
    -   Update vignettes if behavior changes
3.  **Deprecating Functions**
    -   Use `.Deprecated()` function
    -   Maintain old function for at least one version
    -   Update documentation to point to new function
    -   Update vignettes to use new function

## Version Control Guidelines

1.  **Commit Messages**
    -   Start with type: feat, fix, docs, style, refactor, test, chore
    -   Include scope in parentheses
    -   Example: `feat(calc): add new decay function`
2.  **Branches**
    -   `main`: Stable release
    -   `develop`: Development version
    -   `feature/*`: New features
    -   `fix/*`: Bug fixes

## Documentation Updates

When adding or modifying functions: 1. Update relevant vignettes 2. Update README.md if needed 3. Update NAMESPACE 4. Run roxygen2 to update documentation 5. Update NEWS.md with changes

# Complete Function Catalog

+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| Current Name                    | /new name                       | Level       | Subcategory                                          | Notes                                                                                          |
+=================================+=================================+=============+======================================================+================================================================================================+
| **parameter_tuning.R**          | -   depricated                  | \-          | \-                                                   |                                                                                                |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **spread_functions.R**          | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `weighted_spread()`             | → `spread_weighted()`           | Middle      | Spread                                               | General spreading function                                                                     |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.spread_single()`              | → `.spread_core()`              | Low         | Internal                                             | Single value spreading                                                                         |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `spread_access()`               | ✓                               | Middle      | Spread                                               | Access distribution                                                                            |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **compute_choice.R**            | should change to check and main |             |                                                      |                                                                                                |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.compute_choice_core()`        | → `deprec`                      | Low         | Core                                                 | -   will re write a fn - so the computation will be in a main fn while the check is in another |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `compute_choice()`              | → `calc_choice`                 | mid         | mid                                                  | Choice probability calculation                                                                 |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
|                                 | → `.chck_calc_choice`           | Low         | check                                                | validation for calc choice                                                                     |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **compute_huff_weights.R**      | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `compute_huff_weights()`        | → `deprec`                      | Semi-High   | Compute                                              | Huff model weights                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **compute_pmf.R**               | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `compute_pmf()`                 | → `transform_pmf()`             | Middle\*\   | Transform                                            | PMF conversion                                                                                 |
|                                 |                                 | independent |                                                      |                                                                                                |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **decay_functions.R**           | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `compute_weights()`             | → `calc_decay()`                | Middle      | Calc                                                 | Calc decay weight                                                                              |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.gaussian_weights()`           |                                 | Low         | Internal                                             | Gaussian decay                                                                                 |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.exponential_weights()`        |                                 | Low         | Internal                                             | Exponential decay                                                                              |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.power_weights()`              |                                 | Low         | Internal                                             | Power decay                                                                                    |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.inverse_weights()`            |                                 | Low         | Internal                                             | Inverse decay                                                                                  |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.binary_weights()`             |                                 | Low         | Internal                                             | Binary decay                                                                                   |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **gather_functions.R**          | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `weighted_gather()`             | → `gather_weighted()`           | Middle      | Gather                                               | General gathering                                                                              |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `gather_demand()`               | ✓                               | Middle      | Gather                                               | Demand aggregation                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **iterative_chain.R**           | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `preprocess_facilities()`       | → `.help_prep_facilities()`     | Low         | Helper                                               | Facility data preparation                                                                      |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `reintegrate_facilities()`      | → `.help_add_0facilities()`     | Low         | Helper                                               | Facility reintegration                                                                         |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `update_facility_names()`       | → depricated                    | Low         | Helper                                               | Name updating                                                                                  |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `run_iterative_core_fast()`     | → `compute_iterative_fast()`    | semi-high\  | Core                                                 | Fast iteration implementation                                                                  |
|                                 |                                 | \*snap      |                                                      |                                                                                                |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `run_iterative_core()`          | → `compute_iterative()`         | semi-high   | Core                                                 | Full iteration implementation                                                                  |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `spax_ifca()`                   |                                 | High        | Main                                                 | Iterative FCA analysis                                                                         |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **measure_access.R**            | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `measure_access()`              | → `compute_access()`            | Semi-High   | Compute                                              | General accessibility                                                                          |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `compute_2sfca()`               | → `spax_e2sfca()`               | High        | Compute                                              | 2SFCA implementation                                                                           |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **normalize_functions.R**       | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.normalize_core()`             | → `depricate`                   | Low         | Core - will move to validate + main than main + core | Core normalization                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `normalize_weights()`           | → `calc_normalize()`            | Middle      | Calculate                                            | Weight normalization                                                                           |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
|                                 | → `.chck_calc_normalize()`      | low         | checker                                              |                                                                                                |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| **sample_pmf.R**                | \-                              | \-          | \-                                                   | \-                                                                                             |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `sample_pmf()`                  | ok                              | semi-high   | Transform                                            | PMF sampling                                                                                   |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.validate_sample_pmf_inputs()` | → `.chck_sample_pmf()`          | Low         | Check                                                | Input validation                                                                               |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.check_and_convert_pmf()`      | → `.ckck_n_convert_pmf()`       | Low         | Check                                                | PMF validation                                                                                 |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.compute_sample_sizes()`       | → `.help_gen_sample_size()`     | Low         | Helper                                               | Sample size computation                                                                        |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.generate_spatial_samples()`   | → .`sample_pmf_core_indp()`     | mid         | Core                                                 | Sample generation                                                                              |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `est_new_samples()`             | → `.help_est_new_sample()`      | Low         | Helper                                               | Sample estimation                                                                              |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+
| `.generate_evolve_samples()`    | → .`sample_pmf_core_evlv()`     | Low         | Core                                                 | Evolution samples                                                                              |
+---------------------------------+---------------------------------+-------------+------------------------------------------------------+------------------------------------------------------------------------------------------------+

## Missing Functions (Suggested Additions)

+-----------------------------+------------+-------------+------------------------------+
| Function Name               | Level      | Subcategory | Purpose                      |
+=============================+============+=============+==============================+
| `spax_2sfca()`              | High       | Main        | Complete 2SFCA workflow      |
+-----------------------------+------------+-------------+------------------------------+
| `spax_montecarlo()`         | High       | Main        | Monte Carlo analysis         |
+-----------------------------+------------+-------------+------------------------------+
| `summarize_accessibility()` | Middle     | Summarize   | Accessibility statistics     |
+-----------------------------+------------+-------------+------------------------------+
| `validate_inputs_check()`   | Low        | Check       | General input validation     |
+-----------------------------+------------+-------------+------------------------------+
| `plot_accessibility()`      | Middle     | Visualize   | Standard accessibility plots |
+-----------------------------+------------+-------------+------------------------------+

Would you like to: 1. Review specific function groups in detail? 2. Discuss the suggested name changes? 3. Prioritize which missing functions to implement first?
