---
title: "Choosing an FCA index or an allocation model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an FCA index or an allocation model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4, dpi = 100)
library(spax)
```

Choose a model by the quantity you need and the observations that can support
it. spax retains its public floating catchment area (FCA) workflow and adds an
experimental corrected static conditional-logit allocation workflow with an
explicit outside option. Existing FCA calls remain valid; they do not need to
be rewritten as allocation fits.

| Question | Starting point | What the output means |
|:--|:--|:--|
| How is weighted supply distributed relative to weighted demand? | `spax_2sfca()`, `spax_e2sfca()`, `spax_m2sfca()` | A model-specific supply-per-demand accessibility index |
| How are request opportunities allocated among facilities and outside contact? | `prepare_allocation()` then `clm_allocation()` | Choice probabilities and expected loads under a declared model |
| Which parameters fit observed utilization or flows? | `fit_allocation()` with an explicit loss and starts | A calibrated mean model plus optimization diagnostics |
| What changes under a supply scenario at the same parameters? | Reprepare and `evaluate_allocation()` | A model-based contrast, not an estimated causal effect |
| What can be predicted at additional locations? | `predict_allocation()` | Requested outputs with probability, demand and capacity validity flags |
| Are facility-count bootstrap intervals available? | `bootstrap_allocation()` after an eligible fit and observation declaration | Conditional pointwise intervals when the supported procedure's gates pass |

An FCA index is not a contact probability. A fitted model is not automatically
identified or empirically validated. A regional interval is not a cellwise map
interval. These distinctions remain even when both workflows use the same
travel data.

## One generated system, two explicit calculations

This six-origin, three-facility system declares demand as weekly requests,
supply as weekly slots and travel as minutes. Its small raster is a display
template without geographic meaning. All values are generated locally.

```{r shared-inputs}
origin_ids <- paste0("o", 1:6)
facility_ids <- c("a", "b", "c")
D <- setNames(c(300, 200, 400, 100, 350, 250), origin_ids)
S <- c(a = 180, b = 250, c = 150)
T <- cbind(a = c(2, 4, 8, 3, 6, 10),
  b = c(8, 3, 5, 9, 4, 7), c = c(11, 8, 2, 10, 7, 3))
rownames(T) <- origin_ids
grid <- terra::rast(nrows = 2, ncols = 3, xmin = 0, xmax = 3,
  ymin = 0, ymax = 2, crs = "EPSG:3857")
demand_raster <- terra::setValues(grid, unname(D))
travel_raster <- terra::rast(grid, nlyrs = length(S))
terra::values(travel_raster) <- T
names(travel_raster) <- facility_ids
```

The FCA call uses unchanged Gaussian weights in both steps. Its existing API
and result class are preserved.

```{r fca}
fca <- spax_e2sfca(demand_raster, rev(S), travel_raster,
  decay_params = list(method = "gaussian", sigma = 5),
  demand_normalize = "identity")
summary(fca)
plot(fca, main = "FCA supply-per-demand index")
```

The static CLM instead forms a facility score S_j times its decay weight and
divides by the sum of scores plus the outside weight v0. Each row includes an
outside probability, so the inside probabilities can sum to less than 1. Supply
affects attraction here; it is not a hard utilization ceiling.

```{r allocation}
metadata <- list(demand_units = "requests", demand_period = "week",
  supply_units = "slots", supply_period = "week", travel_units = "minutes")
inputs <- prepare_allocation(D, S, T, metadata = metadata)
model <- clm_allocation(inputs, family = "gaussian", fit_v0 = TRUE)
theta <- c(sigma = 5, v0 = 50)
fixed <- evaluate_allocation(model, theta)
prediction <- predict_allocation(fixed, T, demand = D,
  outputs = c("rho", "A", "abar", "flow"))
data.frame(origin_id = origin_ids,
  FCA_index = terra::values(fca$accessibility, mat = FALSE),
  CLM_contact = prediction$outputs$rho,
  CLM_attribution = prediction$outputs$A)
as.data.frame(fixed$coverage)
```

Here `rho` is inside-contact probability. `A` is proportional capacity
attribution per request opportunity, and `abar = A/rho` is attribution per
inside contact where defined. `flow = D * P` is expected request count, and
facility utilization is its column sum. Capacity attribution is not delivered
service. The table places distinct measures beside one another; it is not an
equivalence check or a ranking of model quality.

## Calibrate only after declaring the target

For illustration, round this model's expected facility totals to generate a
target. These counts are synthetic and supply no evidence that the model is
correct for an application. The observed vector is deliberately permuted; names
bind it to the canonical facility IDs.

```{r small-fit}
observed <- setNames(round(colSums(prediction$outputs$flow)), facility_ids)
starts <- rbind(near = c(sigma = 4, v0 = 40), far = c(sigma = 8, v0 = 80))
fit <- fit_allocation(model, rev(observed), starts,
  lower = c(sigma = 1, v0 = 1), upper = c(sigma = 20, v0 = 200),
  output = "utilization", loss = "poisson", gradient = TRUE)
stopifnot(identical(fit$status, "selected"))
as.data.frame(fit, type = "parameters")
summary(fit)
```

The six experimental verbs are **prepare, declare, fit, evaluate, predict and
bootstrap**. They are separate steps so that a chosen fixed model can be
evaluated without fitting, while an eligible fit can be inspected without
refitting. An outside option does not automatically represent unmet need, and
Poisson fitting does not impose exact total matching.

This tiny example is not an uncertainty demonstration. Continue to the
[full allocation workflow](allocation-workflow.html) for the generated
80-origin, 40-facility example with all 399 bootstrap attempts, at least 380
finite successful values per reported quantity, paired supply scenarios and
fresh-process saved-result inspection. Observation declarations and original-fit
eligibility remain required; a bootstrap is not available for every fit.

## Carry these checks across workflows

Use stable facility IDs, declare counts versus densities and reference periods,
and check travel geometry and units before computing. Keep zero demand, unknown
demand and absent travel edges distinct. Legacy FCA reductions and experimental
allocation preparation have different missing-data policies; sharing a file
does not prove that their included support is the same.

For details, read [data preparation](spax-102-data-prep.html),
[the FCA calculation](spax-103-dissection.html),
[resolution and cost](spax-104-raster-tradeoff.html), and
[saved-result inspection](function-naming.html).
