---
title: "2. Preparing counts, facility IDs and travel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Preparing counts, facility IDs and travel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4, dpi = 100)
library(spax)
```

Both FCA and allocation models need a declared demand population, a facility
registry and travel between them. Common preparation principles do not make
their missing-value policies or output meanings identical. This generated
example runs entirely offline. It uses synthetic projected coordinates; no
polygons, facility records or travel services are downloaded.

## Convert area totals to cell counts

Suppose two areas contain 480 and 720 weekly request opportunities. The polygons
align with the generated grid so that each cell belongs to exactly one area.
We distribute each total in proportion to cell area, then check the totals.
This is an allocation assumption, not observed within-area demand.

```{r area-counts}
template <- terra::rast(nrows = 6, ncols = 8, xmin = 0, xmax = 8000,
  ymin = 0, ymax = 6000, crs = "EPSG:3857")
box <- function(left, right) sf::st_polygon(list(rbind(
  c(left, 0), c(right, 0), c(right, 6000), c(left, 6000), c(left, 0))))
zones <- sf::st_sf(zone_id = 1:2, requests = c(480, 720),
  geometry = sf::st_sfc(box(0, 4000), box(4000, 8000), crs = 3857))
zone_grid <- terra::rasterize(terra::vect(zones), template, field = "zone_id")
zone_id <- terra::values(zone_grid, mat = FALSE)
area_km2 <- terra::values(terra::cellSize(template, unit = "km"), mat = FALSE)
counts <- numeric(terra::ncell(template))
for (j in seq_len(nrow(zones))) {
  cells <- which(zone_id == zones$zone_id[j])
  counts[cells] <- zones$requests[j] * area_km2[cells] / sum(area_km2[cells])
}
demand <- terra::setValues(template, counts)
names(demand) <- "weekly_requests"
check <- data.frame(zone_id = zones$zone_id, original = zones$requests,
  reconstructed = as.numeric(tapply(counts, zone_id, sum)))
stopifnot(max(abs(check$original - check$reconstructed)) < 1e-10)
check
terra::plot(demand, main = "Generated request counts per cell", axes = FALSE)
```

If a source raster is a density, multiply it by cell area in the matching area
unit before passing it as demand. Summing densities does not conserve counts.
Real administrative boundaries often cross cells: document fractional overlap,
coastline masks and any renormalization instead of assuming a centroid-based
rasterization preserves totals. A suitable projected CRS helps with local
geometry; a CRS label alone does not guarantee equal physical cell areas.

## Keep facility identity separate from order

```{r facility-travel}
facilities <- data.frame(id = c("f1", "f2", "f3"),
  x = c(1000, 4500, 7000), y = c(1000, 5000, 2000),
  weekly_slots = c(150, 250, 200))
origin_ids <- paste0("origin", seq_len(terra::ncell(template)))
xy <- terra::xyFromCell(template, seq_len(terra::ncell(template)))
travel <- vapply(seq_len(nrow(facilities)), function(j) {
  sqrt((xy[, 1] - facilities$x[j])^2 + (xy[, 2] - facilities$y[j])^2) / 500
}, numeric(nrow(xy)))
dimnames(travel) <- list(origin_ids, facilities$id)
travel_raster <- terra::rast(template, nlyrs = ncol(travel))
terra::values(travel_raster) <- travel
names(travel_raster) <- colnames(travel)
stopifnot(!anyDuplicated(facilities$id), !anyDuplicated(origin_ids),
  terra::compareGeom(demand, travel_raster),
  all(is.finite(counts)), all(counts >= 0),
  all(is.finite(travel)), all(travel >= 0))
```

These travel values assume a constant 500 metres per minute along straight
lines. This deterministic fixture is useful for learning and tests; it is not
a substitute for road-network travel. The origin registry identifies rows;
raster cell numbers identify positions within one particular template.

## Optional live routing, independent of the example

To obtain actual routing inputs, use a suitable routing service and record the
network version, mode, direction and units. Origin-to-facility travel need not
equal facility-to-origin travel on a directed road network. The following
illustration is deliberately not executed during a build and is not used by
any later code.

```{r optional-live-routing, eval=FALSE}
# Supply real facility coordinates and configure your routing service first.
point_wgs84 <- sf::st_transform(real_facilities[1, ], 4326)
live_isochrones <- osrm::osrmIsochrone(
  loc = as.numeric(sf::st_coordinates(point_wgs84)),
  breaks = c(5, 10, 15), res = 30)
```

For a runnable illustration of the geometry-to-raster step, create nested
straight-line catchments around the first synthetic facility. Assign each cell
to the smallest containing catchment. These are illustrative distance rings,
not OSRM output. Cells outside the largest ring remain unknown in this coarse
travel representation.

```{r deterministic-rings}
point <- sf::st_as_sf(facilities[1, ], coords = c("x", "y"), crs = 3857)
breaks <- c(5, 10, 15)
ring_minutes <- rep(NA_real_, terra::ncell(template))
for (limit in rev(breaks)) {
  buffer <- sf::st_buffer(point, dist = limit * 500)
  inside <- terra::values(terra::rasterize(terra::vect(buffer), template,
    field = 1, background = NA), mat = FALSE)
  ring_minutes[!is.na(inside)] <- limit
}
illustrative_rings <- terra::setValues(template, ring_minutes)
terra::plot(illustrative_rings, main = "Illustrative upper travel bands (minutes)", axes = FALSE)
```

Do not add nested polygon values together: that would count overlapping bands
more than once. A real isochrone source may contain disjoint bands or nested
polygons; inspect its definition before rasterization.

## Declare support and missingness

The legacy FCA workflow accepts raster inputs and has its own reduction rules;
for example, `gather_weighted()` defaults to `na.rm = TRUE`. An unknown value
can therefore be omitted from a denominator. It is not automatically treated
as an explicit model-level unknown-data diagnosis. Audit the analysis mask,
excluded demand and travel availability before using that path.

The experimental numeric allocation preparation has explicit policies and a
support registry. Known zero demand, unknown demand and a known absent travel
edge are different declarations. Here zero demand has no fitted load; an
excluded unknown origin has no fitted load because of the stated policy.

```{r checked-allocation-inputs}
D <- setNames(counts, origin_ids)
D[1] <- 0
D[2] <- NA_real_
S <- setNames(facilities$weekly_slots, facilities$id)
units <- list(demand_units = "requests", demand_period = "week",
  supply_units = "slots", supply_period = "week", travel_units = "minutes")
prepared <- prepare_allocation(D, rev(S), travel[, c(3, 1, 2)],
  missing_demand = "exclude", metadata = units)
model <- clm_allocation(prepared, fit_v0 = TRUE)
evaluated <- evaluate_allocation(model, c(sigma = 5, v0 = 50))
requested <- predict_allocation(evaluated, travel, demand = D,
  outputs = c("rho", "flow"))
head(requested$validity)
stopifnot(all(requested$outputs$flow[1, ] == 0),
  all(is.na(requested$outputs$flow[2, ])))
```

With finite travel, both origins can have valid contact probabilities even
though one flow is zero and the other is unavailable. Strict numeric preparation
rejects unknown active travel by default. Positive infinity declares an absent
edge; `NA` declares unknown travel and must not be silently relabeled as an
absent edge. Raster preparation retains its documented legacy travel policy.
Check the selected route's help before comparing raster and numeric inputs.

The unit declarations are recorded, not converted. Prediction travel and demand
must use the saved units and reference period. See [Choosing a model](model-choice.html)
and the [allocation workflow](allocation-workflow.html) for result meanings.
