---
title: "1. Getting started with an FCA accessibility surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting started with an FCA accessibility surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4, dpi = 100)
library(spax)
```

Floating catchment area (FCA) methods combine demand, facility supply and travel
weights to produce an accessibility index. This example uses the existing
`spax_e2sfca()` interface and a small generated grid. It needs no download,
routing server or research dataset. For a fitted allocation model with an
outside option, start with [Choosing a model](model-choice.html).

## Declare the inputs and units

Demand is a **count per cell**, not a density per square kilometre. Supply is
weekly service slots. Distances below are illustrative travel minutes computed
from straight-line separation at a constant 500 metres per minute; they are not
road-network estimates. The projected grid is synthetic and has no geographic
interpretation.

```{r generated-inputs}
demand <- terra::rast(nrows = 6, ncols = 8, xmin = 0, xmax = 8000,
  ymin = 0, ymax = 6000, crs = "EPSG:3857")
terra::values(demand) <- 20 + (seq_len(terra::ncell(demand)) %% 7) * 5
names(demand) <- "weekly_requests"
xy <- terra::xyFromCell(demand, seq_len(terra::ncell(demand)))
facilities <- data.frame(id = c("clinic_a", "clinic_b", "clinic_c"),
  x = c(1500, 4500, 7000), y = c(1000, 5000, 2500),
  slots = c(250, 300, 200))
travel_values <- vapply(seq_len(nrow(facilities)), function(j) {
  sqrt((xy[, 1] - facilities$x[j])^2 + (xy[, 2] - facilities$y[j])^2) / 500
}, numeric(nrow(xy)))
travel <- terra::rast(demand, nlyrs = nrow(facilities))
terra::values(travel) <- travel_values
names(travel) <- facilities$id
stopifnot(terra::compareGeom(demand, travel),
  identical(names(travel), facilities$id))
```

Each travel layer names its facility. Supply can be reordered when it is a data
frame with explicit IDs; an unnamed supply vector or matrix requires the caller
to maintain the correct facility order. Keep an external record of the demand,
supply and travel units: the legacy FCA result does not have the allocation
workflow's structured unit declarations.

## Calculate and inspect the surface

We explicitly choose `demand_normalize = "identity"`: the distance weights are
used unchanged in the demand denominator. With Gaussian scale 5 minutes, a
distance of 5 minutes has weight `exp(-1/2)`, not zero.

```{r calculate}
result <- spax_e2sfca(demand, facilities[c(3, 1, 2), ], travel,
  decay_params = list(method = "gaussian", sigma = 5),
  demand_normalize = "identity", id_col = "id", supply_cols = "slots")
print(result)
summary(result)
plot(result, main = "FCA index: weekly slots per weekly request")
```

The result is a `spax` object. Its `accessibility` member is a `SpatRaster`;
`facilities` records supply and `parameters` records the decay and normalization
choices. This index has supply-per-demand units. It is not a contact probability,
an observed service rate, or the CLM contact measure `rho`.

```{r inspect-and-check}
head(terra::as.data.frame(result$accessibility, xy = TRUE, na.rm = FALSE))
result$parameters
# Independent check of the declared identity-normalized FCA calculation.
W <- exp(-travel_values^2 / (2 * 5^2))
D <- terra::values(demand, mat = FALSE)
load <- colSums(W * D)
expected <- drop(W %*% (facilities$slots / load))
stopifnot(isTRUE(all.equal(
  terra::values(result$accessibility, mat = FALSE), expected,
  check.attributes = FALSE, tolerance = 1e-10)))
```

The spatial summary describes raster cells; it is not automatically weighted by
demand. If your question concerns the average request, calculate and label that
weighted summary explicitly.

```{r weighted-summary}
data.frame(request_weighted_index = weighted.mean(expected, D),
  total_slots_per_request = sum(facilities$slots) / sum(D))
```

The equality here follows from using the same weights on both sides with
identity normalization, positive facility loads and this fully observed domain.
Do not transfer it to other normalizations or incomplete support without
checking their equations.

## Continue

- [Prepare counts, IDs and travel](spax-102-data-prep.html), including unknown
  values and an explicitly optional live-routing example.
- [Inspect the FCA calculation](spax-103-dissection.html) and its alternatives.
- [Choose resolution and output size](spax-104-raster-tradeoff.html).
- [Fit a static allocation model](allocation-workflow.html) when expected
  utilization, explicit outside contact and supported uncertainty are required.
