---
title: "4. Resolution, support and computational cost"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Resolution, support and computational cost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4, dpi = 100)
library(spax)
```

Rasters provide explicit geometry and alignable cell support. They also incur
costs that grow with origin count, facility count and requested outputs. More
cells do not by themselves improve travel or demand information. The generated
example below measures changes in representation, not hardware performance.
It uses no large bundled datasets or optional benchmarking package.

## What a dense origin-by-facility array costs

```{r storage}
dimensions <- data.frame(origins = c(1000, 120000, 1000000),
  facilities = c(10, 40, 100))
dimensions$one_double_array_MiB <- with(dimensions, origins * facilities * 8 / 1024^2)
dimensions
```

This is only the numeric payload of one dense double array. A workflow can hold
travel, kernels, temporary arrays and optional flows at the same time. Raster
object size may reflect an external pointer or a file reference rather than
all resident data. Use process-level peak memory and elapsed time for an actual
workflow; do not infer either from `object.size()` or a single tiny operation.

## Aggregate counts, then state what happened to travel

```{r generated-grid}
fine <- terra::rast(nrows = 12, ncols = 16, xmin = 0, xmax = 8000,
  ymin = 0, ymax = 6000, crs = "EPSG:3857")
terra::values(fine) <- 10 + (seq_len(terra::ncell(fine)) %% 11)
xy <- terra::xyFromCell(fine, seq_len(terra::ncell(fine)))
sites <- data.frame(id = c("a", "b", "c"), x = c(1000, 4500, 7000),
  y = c(1000, 5000, 2000), slots = c(150, 250, 200))
travel <- terra::rast(fine, nlyrs = 3)
terra::values(travel) <- vapply(1:3, function(j) {
  sqrt((xy[, 1] - sites$x[j])^2 + (xy[, 2] - sites$y[j])^2) / 500
}, numeric(nrow(xy)))
names(travel) <- sites$id
```

We sum demand counts when coarsening cells. For this demonstration only, we
average fine-cell travel values in each larger cell. Averaging travel before a
nonlinear decay transformation does not generally preserve averaged weights or
accessibility. It is one explicit approximation, not a recommended default for
every network.

```{r resolution-comparison}
factors <- c(1, 2, 4)
answers <- lapply(factors, function(factor) {
  D <- if (factor == 1) fine else terra::aggregate(fine, fact = factor, fun = sum)
  T <- if (factor == 1) travel else terra::aggregate(travel, fact = factor, fun = mean)
  result <- spax_e2sfca(D, sites, T,
    decay_params = list(method = "gaussian", sigma = 5),
    demand_normalize = "identity", id_col = "id", supply_cols = "slots")
  list(demand = D, result = result)
})
comparison <- do.call(rbind, lapply(seq_along(answers), function(k) {
  D <- terra::values(answers[[k]]$demand, mat = FALSE)
  A <- terra::values(answers[[k]]$result$accessibility, mat = FALSE)
  data.frame(factor = factors[k], cells = length(D), demand = sum(D),
    minimum = min(A), maximum = max(A), request_weighted_mean = weighted.mean(A, D))
}))
stopifnot(max(abs(comparison$demand - sum(terra::values(fine)))) < 1e-10)
comparison
```

Total demand is preserved, but local patterns need not be. Here the weighted
mean remains total supported supply divided by total demand because both FCA
steps use the same identity-normalized kernel with positive loads. That
identity does not establish local accuracy. Other normalizations and travel
support can change the accounting relationship.

```{r coarse-map}
plot(answers[[3]]$result, main = "Illustrative FCA after count aggregation")
```

## Preparation and requested outputs matter

The experimental allocation workflow prepares the active load system once and
reuses it for multiple starts. Zero and excluded unknown demand do not become
positive fitted demand. Requested-location prediction is a separate operation;
it can evaluate contact probabilities outside fitted load support under its
declared validity rules.

Request only needed outputs. A regional table or an origin vector is smaller
than a full origin-by-facility flow matrix. `predict_allocation(...,
outputs = "rho")` requests contact; adding `"flow"` asks for expected flows and
requires requested demand. Creating a map also requires an explicit spatial
mapping. Saving small prediction snapshots does not mean all result objects or
legacy raster objects are pointer-free; use their documented persistence path.

For a deployment-sized acceptance test, record the exact input dimensions,
active support, requested outputs, extraction cost, runtime, peak resident
memory, persistence and numerical agreement before changing the implementation.
Do not compare runs with different output retention or overlapping heavy jobs.
The [full allocation vignette](allocation-workflow.html) deliberately retains
all 399 bootstrap attempts; it is a more expensive validation example, not a
microbenchmark to repeat while measuring another task.
