---
title: "3. Understanding the FCA calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Understanding the FCA calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
  fig.height = 4, dpi = 100)
library(spax)
```

An FCA calculation has two kernels: one determines weighted demand at each
facility, and the other returns each facility's supply-to-demand ratio to
origins. Let demand be D_i, supply be S_j, demand weights be G_ij and access
weights be H_ij. The calculation is

\[
L_j=\sum_i D_iG_{ij},\qquad R_j=S_j/L_j,\qquad A_i=\sum_j H_{ij}R_j.
\]

The public FCA implementation uses zero for the facility ratio when its weighted
demand denominator is zero. That is a computational convention, not evidence
that unused capacity is physically absent. This example has positive loads and
fully observed inputs, so it does not invoke that convention.

## A small network with explicit IDs

```{r inputs}
demand <- terra::rast(nrows = 2, ncols = 3, xmin = 0, xmax = 3,
  ymin = 0, ymax = 2, crs = "EPSG:3857")
terra::values(demand) <- c(30, 20, 40, 10, 35, 25)
distance <- terra::rast(demand, nlyrs = 3)
travel <- cbind(a = c(2, 4, 8, 3, 6, 10),
  b = c(8, 3, 5, 9, 4, 7), c = c(11, 8, 2, 10, 7, 3))
terra::values(distance) <- travel
names(distance) <- colnames(travel)
supply <- c(a = 18, b = 25, c = 15) # weekly slots
```

Travel is minutes, demand is weekly requests and supply is weekly slots. The
grid is only a display template; its coordinates did not generate these travel
times. A travel layer's facility ID must agree with the corresponding supply ID.

## Decay and normalization are separate choices

For Gaussian decay, W_ij = exp(-d_ij^2/(2 sigma^2)). The same sigma has the travel
unit; it is neither a catchment boundary nor a fitted parameter in this example.

```{r kernels}
weights <- calc_decay(distance, method = "gaussian", sigma = 5)
standard <- calc_normalize(weights, method = "standard")
semi <- calc_normalize(weights, method = "semi")
data.frame(cell = seq_len(terra::ncell(demand)),
  raw_sum = rowSums(terra::values(weights)),
  standard_sum = rowSums(terra::values(standard)),
  semi_sum = rowSums(terra::values(semi)))
```

For raster weights and no outside weight, `identity` leaves the kernel unchanged;
`standard` divides each cell's weights by their sum when that sum is positive;
`semi` divides only when the sum exceeds 1. These choices change G and therefore
the denominator L. Normalization is a model choice, not a universal correction
for an error. In `spax_e2sfca()`, `demand_normalize` acts on the demand-side
weights; the access-side weights remain W.

## Compose the public operations

Here G=H=W and normalization is explicitly `identity`. The named output of
`gather_weighted()` is aligned before spreading: `spread_weighted()` uses
positional values and does not perform an ID join for us.

```{r compose}
weighted_demand <- gather_weighted(demand, weights, simplify = TRUE)
ratio <- supply[names(weighted_demand)] / weighted_demand
access <- spread_weighted(unname(ratio[names(weights)]), weights)
wrapped <- spax_e2sfca(demand, rev(supply), distance,
  decay_params = list(method = "gaussian", sigma = 5),
  demand_normalize = "identity")
custom <- compute_access(demand, supply, weights, weights)
W <- exp(-travel^2 / (2 * 5^2))
D <- terra::values(demand, mat = FALSE)
manual <- drop(W %*% (supply / colSums(W * D)))
for (answer in list(access, wrapped$accessibility, custom)) {
  stopifnot(isTRUE(all.equal(terra::values(answer, mat = FALSE), manual,
    tolerance = 1e-10, check.attributes = FALSE)))
}
plot(wrapped, main = "Identity-normalized FCA index")
```

The wrapper, public component path and independent matrix calculation agree for
these declared inputs. The check verifies this calculation; it does not imply
that every accessibility model is mathematically equivalent.

## Change the model explicitly

`spax_2sfca()` uses a binary travel threshold. `spax_m2sfca()` uses W on the
demand side and W squared on the access side. Neither change is a fitted
statistical comparison.

```{r alternatives}
binary <- spax_2sfca(demand, supply, distance, threshold = 6)
modified <- spax_m2sfca(demand, supply, distance,
  decay_params = list(method = "gaussian", sigma = 5),
  demand_normalize = "identity")
modified_manual <- drop((W^2) %*% (supply / colSums(W * D)))
stopifnot(isTRUE(all.equal(terra::values(modified$accessibility, mat = FALSE),
  modified_manual, tolerance = 1e-10, check.attributes = FALSE)))
data.frame(cell = seq_along(D), E2SFCA = manual,
  M2SFCA = modified_manual, binary_FCA = terra::values(binary$accessibility, mat = FALSE))
```

In a corrected static CLM, supply-weighted attraction and an outside option
instead define choice probabilities. Expected facility loads and contact
probabilities then have explicit allocation meanings. An FCA index must not be
relabeled as a CLM probability. [Choosing a model](model-choice.html) explains
the distinction; [the allocation walkthrough](allocation-workflow.html) shows
fitting and checking the separate model.
